Query Optimization for Dynamic Imputation

### 2017-05-12 paper/algorithm.tex
paper: pass over imputation methods section

\ProjectName{} is designed so that any imputation strategy
can be plugged in with minimal effort and without changes to the optimizer, simply by supplying
penalty ($P$) and time ($T$) functions (\Cref{sec:cost-model}) pertinent to the algorithm.
These can then be queried by the optimizer during planning.  The flexibility of
this approach aligns with the spirit of query optimization, which aims to
reconcile performance with a declarative interface to data manipulation. To
motivate this API, we describe how an analyst would specify the form of these
functions for a general-purpose imputation algorithm, as well as two simpler
methods.

@@

As a reference implementation,
\ProjectName{} uses a general-purpose imputation strategy based on
chained-equation decision trees.  Chained-equation imputation
methods~\cite{vanbuuren2011mice} (sometimes called \textit{iterative
regression}~\cite{gelman2006data}) impute multiple missing attributes by
iteratively fitting a predictive model of one missing attribute, conditional on
all complete attributes and all other missing attributes. Chained-equation
decision trees algorithms are found to be effective~\cite{akande2015empirical}
and are widely used in epidemiological domains~\cite{burgette2010multiple}.

@@

The chained-equation algorithm proceeds by
iteratively fitting decision trees for the target imputation attributes. In
each iteration, the missing values of a separate attribute are replaced with
newly imputed values.  Values in a single column can be updated over multiple
iterations of the algorithm. With more iterations, the quality of the
imputation improves as values progressively reflect more accurate relationships
between attributes. The algorithm terminates after a fixed number of
iterations, or earlier if convergence is achieved, indicated by unchanged values in a 
column in subsequent iterations. For our experiments, we use Weka's \texttt{REPTree}
implementation~\cite{witten2016data}, a variant of the canonical \texttt{C4.5}
algorithm~\cite{quinlan1993}. Roughly, the time complexity of building one
decision tree is $\mathcal{O}(nm \text{ } log(n))$, where $n$ is the number of tuples
and $m$ is the number of dependent attributes~\cite{witten2016data}. Since we build
$k * |\textsc{Dirty}(q)|$ trees, we have

@@

Separately, we specify a heuristic penalty that takes into account the
improvement in imputation quality from more tuples and attributes as well as
the diminishing returns to more data.

@@

At the other end of the spectrum of complexity, an analyst could use a simple
method that replaces a missing value with the mean over available values
in that attribute, a common choice in exploratory data analysis known
as mean value imputation.

We can take advantage of the histograms maintained by \ProjectName{}
to estimate the mean and construct a non-blocking impute operator.
Applying the heuristics from \ref{sec:quality} and \ref{sec:query-runtime}, we
have

@@

@@

Hot-deck imputation is a nonparametric method in which missing values are
replaced with randomly selected complete values. This method preserves the
distribution of the complete values. The heuristic $P$ remains the same, 
but we modify $T$ to account for the fact that hot-deck is blocking in our
implementation.

@@

%\begin{algorithm}
%    \input{algorithms/cart.tex}
%\end{algorithm}

>>>>

\ProjectName{} is designed so that any imputation strategy can be plugged in with minimal effort and without changes to the optimizer, by supplying penalty ($P$) and time ($T$) functions (\Cref{sec:cost-model}) pertinent to the algorithm.
These cost functions are then used by the optimizer during planning.
The flexibility of this approach is in the spirit of query optimization, which aims to reconcile performance with a declarative interface to data manipulation.
To motivate this API, we describe how a developer would specify these functions for a general-purpose imputation algorithm, as well as for two simpler imputation methods.

@@

As a reference implementation,
\ProjectName{} uses a general-purpose imputation strategy based on
chained-equation decision trees.  Chained-equation imputation
methods~\cite{vanbuuren2011mice} (sometimes called \textit{iterative
regression}~\cite{gelman2006data}) impute multiple missing attributes by
iteratively fitting a predictive model of one missing attribute, conditional on
all complete attributes and all other missing attributes. Chained-equation
decision trees algorithms are effective~\cite{akande2015empirical}
and are widely used in epidemiological domains~\cite{burgette2010multiple}.

@@

The chained-equation algorithm proceeds by
iteratively fitting decision trees for the target attributes. 
In each iteration, the missing values of a single attribute are replaced with newly imputed values.  Values in a single column can be updated over multiple
iterations of the algorithm. With more iterations, the quality of the
imputation improves as values progressively reflect more accurate relationships
between attributes. The algorithm terminates after a fixed number of
iterations, or earlier if convergence is achieved, indicated by unchanged values in a 
column in subsequent iterations. For our experiments, we use Weka's \texttt{REPTree}
implementation~\cite{witten2016data}, a variant of the canonical \texttt{C4.5}
algorithm~\cite{quinlan1993}. Roughly, the time complexity of building one
decision tree is $\mathcal{O}(nm \text{ } log(n))$, where $n$ is the number of tuples
and $m$ is the number of dependent attributes~\cite{witten2016data}. Since we build
$k * |\textsc{Dirty}(q)|$ trees, we have

@@

Separately, we specify a heuristic penalty that takes into account the
improvement in imputation quality from more tuples and attributes as well as
the diminishing returns from more data.

@@

At the other end of the complexity spectrum, mean value imputation is a common choice for exploratory data analysis.
Mean value imputation replaces missing values in a column with the mean of the available values in the column.

We can take advantage of the histograms maintained by \ProjectName{} to estimate the mean and construct a non-blocking impute operator.
Applying the heuristics from~\Cref{sec:quality} and~\Cref{sec:query-runtime}, we
have

@@

Note that $P$ increases with the variance of the columns and decreases with their cardinality, reflecting the intuition that the quality of the mean is lower with high variance and higher with more data.

@@

Hot-deck imputation is a nonparametric method in which missing values are replaced with randomly selected complete values.
This method preserves the distribution of the complete values.
We use the same $P$ as for mean value imputation, but we modify $T$ to account for the fact that hot-deck is blocking in our implementation.

@@
